---
title: "Document Classification"
author: "Desiree Thomas, Denise Atherley"
format: html
editor: visual
---

## Introduction

Almost everyone who has likely received a spam email (whether in the spam folder or in their Inbox) or had a legitimate email flagged as spam, where it was banished to the Spam Folder. In my own personal experience, Outlook has consistently targeted my NYC Comic Con verification as spam, while allowing fake job offers to enter my inbox. It can be very frustrating to users. In this assignment, we attempted document classification, where we were given spam and ham test documents from the spamassassin corpus. The goal for this assignment was to train a model to recognize the emails as either spam or ham. Please note that due to having to download, unzip and store the test documents, there was no way for us to have a RAW GitHub url that available for use. When running this file, please alter the file paths to match your own. Here is the url for the test documemts corpus: https://spamassassin.apache.org/old/publiccorpus/



## Load in libraries 
```{r}
library(tidyverse)
library(tidytext)
library(tm)
library(e1071)
library(caret)
```

```{r}
#Ingestion 

read_corpus <- function(directory, class_label) {
  files <- list.files(directory, full.names =  TRUE)
  
  #map_df with encoding to prevent UTF-8 errors 
  df <- map_df(files, ~tibble(
    text = read_file(.x, locale = locale(encoding = "latin1")), 
    label = class_label, 
    file_name = basename(.x) 
  ))
  return(df)
}

```


IMPORTANT: Alter file paths to match your own for reproducibility.
```{r}
#Paths are based on working directory; adjust for your own working directory
ham_data <- read_corpus("/Users/dthom/Desktop/Data_607/document-classification/Document-Classification/project4/data/raw/easy_ham", "ham")
spam_data <- read_corpus("/Users/dthom/Desktop/Data_607/document-classification/Document-Classification/project4/data/raw/spam", "spam")

#Combine and create a document ID 
email_corpus <- bind_rows(ham_data, spam_data) %>% 
  mutate(doc_id = row_number())

```


## Cleaning and vectorization

In this chunk, we used this as our cleaning phase to remove noise and vectorization to create the DTM. The tidytext package was very helpful for this portion. Here, we tokenized and cleaned by breaking the text into words and removed standard stop words. 
```{r}
# Force R to load the stop words dataframe due to error in previous iteration
data("stop_words", package = "tidytext")

# Runs the pipeline responsible for cleaning
tidy_emails <- email_corpus %>%
  unnest_tokens(word, text) %>%
  anti_join(tidytext::stop_words, by = "word") %>%
  filter(!stringr::str_detect(word, "^[0-9]+$")) %>%
  filter(nchar(word) > 2)
```

This counts the words per document and casts it to the matrix that our modeling algorithm needs. 
```{r}
#Creates the DTM
email_dtm <- tidy_emails %>%
  count(doc_id, word) %>%
  cast_dtm(doc_id, word, n)

#Keeps words that appear in at least 1% of emails 
email_dtm_reduced <- tm::removeSparseTerms(email_dtm, 0.99)

#Converts to a dataframe that the algorithm can actually read
dtm_df <- as.data.frame(as.matrix(email_dtm_reduced))

#Adds the spam and ham labels to the new matrix
dtm_df$label <- email_corpus$label[match(rownames(dtm_df), email_corpus$doc_id)]
dtm_df$label <- as.factor(dtm_df$label)
```

# Training and evaluation 
 
We used 80% of our emails to train the model and the other 20% were used to test the model. We used the Naive Bayes as our model of choice and caret was used for the Confusion Matrix. We set a seed, to help aid reproducibility.

```{r}

# Creates a function that converts any number > 0 to "Yes", and 0 to "No"
convert_counts <- function(x) {
  factor(ifelse(x > 0, "Yes", "No"))
}

# Identifies all columns EXCEPT the 'label' column
cols_to_convert <- setdiff(names(dtm_df), "label")

# Applies the function to the word columns to overwrite the numbers with Yes/No
dtm_df[cols_to_convert] <- lapply(dtm_df[cols_to_convert], convert_counts)

```


```{r}
set.seed(42) #this is our seed 
train_index <- sample(1:nrow(dtm_df), 0.8 * nrow(dtm_df))
train_data <- dtm_df[train_index, ]
test_data <- dtm_df[-train_index, ]


```

#Model Training

```{r}
nb_model <- naiveBayes(label ~ ., data = train_data) #There may be some lag here depending on your machine
#this is using the 20% that was withheld from the model make predictions
predictions <- predict(nb_model, test_data)

#Evaluate
confusionMatrix(predictions, test_data$label)

```
In the first iteration, the model predicted 951 false negatives; it predicted 951 "ham" emails as "spam". This may have been due to the naiveBayes() function being given a dataframe that had a lot of numerical term counts such as "free" a number of times. The algorithm most likely then assumed that the words that came afterward was following a normal distribution. This led the model to try to fit the bell curve to a matrix that had mostly 0's inside of it, which then wrecked the variance, which in turn led the model to put everything into the "spam" section. 

We asked Gemini and it suggested that we use the Bernoulli Naive Bayes. The change was made due to Naive Bayes needing categorical data and not continuous data. The model should be tracking if words are appearing or not, not just the frequency of their appearance. To deal with this, we decided to then enter a conversion step right before we split the data and set the seed. 

After making this change, our Accuracy jumped from 24% to 97% as well as the Kappa score becoming statistically significant. 1019 real emails made it through and only 1 spam email tricked the model, this is the predicted ham. The spam folder actually caught 215 spam emails, though it did accidentally flag 26 of the real emails as spam, this is the predicted ham. 

## Conclusion

Our final model ended up being a Bernoulli Naive Bayes classifier after we saw the accuracy reported after our first run. We converted to this classifier because we needed to convert the raw word counts into categorical "yes and no" factors to handle the sparse matrix correctly. Our model's final accuracy score was 97.86% with only 1 spam email bypassing the filter. The pipeline also ingested the 2003 corpus (I started with only a few to prove that the pipeline worked before I scaled and added the rest of the corpus). 

## Future Iterations 
Due to debugging time, the original plan to experiment with a Support Vector Machine and scraping web pages has been postponed to a future iteration to ensure that the deliverable is...delivered... on time. 

#Citations 
Gemini was used to both provide terminal commands to avoid using the GUI when extracting and making directories, alongside providing debugging support. It also provided an explanation on why the accuracy score was so low and suggested the Bernoulli alternative.

Google DeepMind. (2026). Gemini 3.1 Pro [Large Language model].
https://gemini.google.com. Accessed May 03, 2026
